---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Anahatt Virk"
date: "07/04/2025"
output: github_document
---

### Load packages and data
```{r info}
library(usethis)
use_git_config(
  user.name = "Anahatt Virk",
  user.email = "virka22@wfu.edu")
```

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

### Exercise 1

```{r load-data}
staff <- read_csv("data/instructional-staff.csv")
```

```{r reshape}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

```{r plot}
staff_long %>%
  ggplot(aes(x = year, y = value, group = faculty_type, color = faculty_type)) +
  geom_line() +
  labs(title = "Instructional Staff Over Time by Faculty Type", x = "Year", y = "Number of Staff", color = "Faculty Type") +
  theme_minimal()
```

### Exercise 2

To show that the proportion of part-time faculty have gone up over time, I would change the previous plot in order to show the proportions within each year, instead of the raw counts. The y-axis label would have to be updated to reflect proportions instaed of number of staff. The updated graph has been included below. 

```{r proportions}
staff_prop <- staff_long %>%
  group_by(year) %>%
  mutate(proportion = value / sum(value)) %>%
   ungroup()
```

```{r proportion-plot}
staff_prop %>%
  ggplot(aes(x = year, y = proportion, group = faculty_type, color = faculty_type)) +
  geom_line() +
  labs(title = "Proportion of Instructional Staff by Faculty Type Over Time", x = "Year", y = "Proportion of Staff", color = "Faculty Type") +
   theme_minimal()
```

### Exercise 3

To improve the original plot, I believe that a cleaner and more interpretable format should be used such as a bar graph. Additionally, while the original includes 17 countries in the visualization, it would be best to limit them to the top 10 fish producing countries, and group the rest together as "other." Capture and aquaculture production could also be depicted on the same graph through the use of stacking.

```{r load-fisheries-data}
fisheries <- read_csv("data/fisheries.csv")

fish_long <- fisheries %>%
    slice_max(total, n = 10) %>%

  pivot_longer(cols = c(capture, aquaculture),
             names_to = "production_type",
             values_to = "tons")
```

```{r fisheries-plot}
ggplot(fish_long, aes(x = tons, y = country, fill = production_type)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("capture" = "blue", "aquaculture" = "green"), labels = c("Capture", "Aquaculture")) +
  labs(title = "Top 10 Fishery-Producing Countries in 2016", x = "Production (Tons)", y = "Country", fill = "Production Type") +
  theme_minimal()
```

### MosaicData

```{r mosaic-data}
library(mosaicData)
data(Whickham)
```

### Smokers - Exercise 1

I believe that these data came from an observational study, as they follow smoking behavior and outcomes in women without any intervention from the researchers. It would be unethical to randomly assign individuals to smoke or not for the purpose of a study, further supporting my claim that this was observational and not experimental.

### Smokers - Exercise 2

The data frame includes 1314 observations, each representing one participant.  

### Smokers - Exercise 3

There are 3 variables in the dataset: outcome, smoking status, and age. Outcome and smoking status are categorical variables, while age is a numeric variable. 

```{r outcome-visual}
ggplot(Whickham, aes(x=outcome)) +
  geom_bar(fill="blue") + 
  labs(title = "Survival Outcome", x= "Outcome", y = "Count" )
```

```{r smoker-visual}
ggplot(Whickham, aes(x=smoker)) +
  geom_bar(fill="pink") + 
  labs(title = "Smoking Status", x= "Smoker?", y = "Count")
```

```{r age-visual}
ggplot(Whickham, aes(x=age)) +
  geom_histogram(fill="green") +
  labs(title = "Age Distribution", x ="Age", y ="Count") 
```

### Smokers - Exercise 4

I would expect smokers to have worse health outcomes than non-smokers would. In the case of this dataset, I predict that smokers are more likely to be dead after 20 years in comparison to non-smokers.

### Smokers - Exercise 5

```{r smoking-outcome-visual}
Whickham %>%
  count(smoker, outcome) %>%
  ggplot(aes(x = smoker, y = n, fill = outcome)) + 
  geom_col() + 
  labs(title="Outcomes by Smoking Status", x = "Smoking Status", y ="Count", fill = "Health Outcome")
```

```{r probabilities}
Whickham %>%
  count(smoker, outcome) %>%
  group_by(smoker) %>%
  mutate(probability = n / sum(n))
```

The data does not support my expectations. From the graph we can see that there are more non-smokers dead after 20 years than smokers. Additionally, looking at the conditional probabilities also contradicts my previous statement. The probability for a non-smoker to be dead is 0.314, while it is 0.239 for smokers. 

### Smokers - Exercise 6
### Smokers - Exercise 7

