---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Anahatt Virk"
date: "07/04/2025"
output: github_document
---

### Load packages and data
```{r info}
library(usethis)
use_git_config(
  user.name = "Anahatt Virk",
  user.email = "virka22@wfu.edu")
```

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
```

### Exercise 1

```{r load-data}
staff <- read_csv("data/instructional-staff.csv")
```

```{r reshape}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```

```{r plot}
staff_long %>%
  ggplot(aes(x = year, y = value, group = faculty_type, color = faculty_type)) +
  geom_line() +
  labs(title = "Instructional Staff Over Time by Faculty Type", x = "Year", y = "Number of Staff", color = "Faculty Type") +
  theme_minimal()
```

### Exercise 2

To show that the proportion of part-time faculty have gone up over time, I would change the previous plot in order to show the proportions within each year, instead of the raw counts. The y-axis label would have to be updated to reflect proportions instaed of number of staff. The updated graph has been included below. 

```{r proportions}
staff_prop <- staff_long %>%
  group_by(year) %>%
  mutate(proportion = value / sum(value)) %>%
   ungroup()
```

```{r proportion-plot}
staff_prop %>%
  ggplot(aes(x = year, y = proportion, group = faculty_type, color = faculty_type)) +
  geom_line() +
  labs(title = "Proportion of Instructional Staff by Faculty Type Over Time", x = "Year", y = "Proportion of Staff", color = "Faculty Type") +
   theme_minimal()
```

### Exercise 3

To improve the original plot, I believe that a cleaner and more interpretable format should be used such as a bar graph. Additionally, while the original includes 17 countries in the visualization, it would be best to limit them to the top 10 fish producing countries, and group the rest together as "other." Capture and aquaculture production could also be depicted on the same graph through the use of stacking.

```{r load-fisheries-data}
fisheries <- read_csv("data/fisheries.csv")

fish_long <- fisheries %>%
    slice_max(total, n = 10) %>%

  pivot_longer(cols = c(capture, aquaculture),
             names_to = "production_type",
             values_to = "tons")
```

```{r fisheries-plot}
ggplot(fish_long, aes(x = tons, y = country, fill = production_type)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("capture" = "blue", "aquaculture" = "green"), labels = c("Capture", "Aquaculture")) +
  labs(title = "Top 10 Fishery-Producing Countries in 2016", x = "Production (Tons)", y = "Country", fill = "Production Type") +
  theme_minimal()
```


### Exercise 3
### Exercise 3
### Exercise 3
### Exercise 3
Add exercise headings as needed.
